---
title: "Test propellor exp"
author: "Ole Gunnar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")
```

## Create mock data

```{r}
# Load necessary library
library(mgcv)

# Simulating some mock data
set.seed(123) # For reproducibility
n <- 100 # Number of observations

# Independent variables
air_pressure <- runif(n, min = 0.01, max = 1) # Air pressure from 0.01 to 1 atm
propeller_distance <- runif(n, min = -8, max = 10) # Distance from -8 to +10

# Simulate some nonlinear relationships for the mock response
force <- (2 * exp(-0.05 * (propeller_distance + 3)^2)) * exp(-0.3 * (1 - air_pressure))

# Add some noise to force to simulate measurement or process variability
force <- force + rnorm(n, mean = 0, sd = 0.2)

# Combine into a data frame
data <- data.frame(air_pressure, propeller_distance, force)
data$force <- abs(data$force)
head(data)
```

## Plotting Force vs. Air Pressure

```{r}
library(ggplot2)

# Plotting force vs air pressure
ggplot(data, aes(x = air_pressure, y = force)) +
  geom_point() +
  geom_smooth(method = 'gam', formula = y ~ s(x), method.args = list(family = "gaussian")) +
  theme_minimal() +
  labs(title = "Force vs. Air Pressure", x = "Air Pressure (atm)", y = "Force")

```

## Plotting Force vs. Propeller Distance

```{r}
# Plotting force vs propeller distance
ggplot(data, aes(x = propeller_distance, y = force)) +
  geom_point() +
  geom_smooth(method = 'gam', formula = y ~ s(x), method.args = list(family = "gaussian")) +
  theme_minimal() +
  labs(title = "Force vs. Propeller Distance", x = "Propeller Distance (cm)", y = "Force")

```

```{r}
# Fit a GAM model with splines
model <- gam(force ~ s(air_pressure, k = 20) + s(propeller_distance, k = 20), data = data)

# Summary of the model
summary(model)
```

## Diagnostic Plots of the GAM Model

```{r}
# Plotting the effects of the GAM model
plot(model, pages = 1, all.terms = TRUE, se = TRUE, scale = 0)

```

## Plotting in 3d

```{r}
# #install.packages("plotly", dependencies = true)
# library(htmltools)
# library(plotly)
# 
# # creating a 3d plot
# fig <- plot_ly(data, x = ~propeller_distance, y = ~air_pressure, z = ~force, type = 'scatter3d', mode = 'markers',
#                marker = list(size = 5, color = force, colorscale = 'viridis', opacity = 0.8)) %>%
#   layout(title = "3d plot of force vs. propeller distance and air pressure",
#          scene = list(xaxis = list(title = 'propeller distance (cm)'),
#                       yaxis = list(title = 'air pressure (atm)'),
#                       zaxis = list(title = 'force')))
# 
# # display the plot
# fig


```

## Spline surface

To plot a spline surface from the GAM model over your 3D plot, we need to predict values using the model across a grid that covers the range of **`propeller_distance`** and **`air_pressure`**. This involves creating a new dataset that represents all combinations of these two variables within their observed ranges and then using the **`predict`** function to get the corresponding force values from the GAM model. Finally, we can use **`plotly`** to plot the surface.

```{r}
library(plotly)
# Generate a grid for predictions
predict_grid <- expand.grid(propeller_distance = seq(min(data$propeller_distance), max(data$propeller_distance), length = 50),
                            air_pressure = seq(min(data$air_pressure), max(data$air_pressure), length = 50))

# Predict force values using the GAM model over the grid
predict_grid$force <- predict(model, newdata = predict_grid)

# Convert 'force' predictions into a matrix form
z_matrix <- matrix(predict_grid$force, nrow = length(unique(predict_grid$propeller_distance)), byrow = TRUE)


# Plotting the 3D surface of the GAM prediction with original data points
fig <- plot_ly() %>%
  add_surface(x = unique(predict_grid$propeller_distance), 
              y = unique(predict_grid$air_pressure), 
              z = z_matrix, 
              colorscale = 'Viridis', 
              showscale = FALSE) %>% # Hide color scale for the surface
  add_trace(x = data$propeller_distance, y = data$air_pressure, z = data$force, 
            type = 'scatter3d', mode = 'markers', 
            marker = list(size = 2, color = 'rgba(255, 255, 255, 0.5)', line = list(color = 'black', width = 0.5))) %>%
  layout(title = "Predicted Force from GAM Model with Data Points",
         scene = list(xaxis = list(title = 'Propeller Distance (cm)'),
                      yaxis = list(title = 'Air Pressure (atm)'),
                      zaxis = list(title = 'Force')))

# Display the plot
fig

```
